---
title: "Exercises: R in Action"
author: "Ulrich Matter"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data: Student's GPA

The data used in the exercices are based on .....
The data file is in sav-format (SPSS). We can read data from other stats packages into R with the **foreign** library.
Thus, first, we load and install this R-package:

```{r foreign, message=FALSE, results=FALSE}
install.packages("foreign", repos = 'http://cran.us.r-project.org')
library(foreign)
```

Read the data into R as follows:
```{r data, warning=FALSE}
sample <- read.spss("data/sample_data.sav", to.data.frame = TRUE)
```

Have a looka at the data set:
```{r view}
names(sample)
View(sample)
```

"Clean" the data
```{r clean}
sample <- sample[!is.na(sample$SSATScore),] # remove observations without SSATScore
sample <- sample[!is.na(sample$HSGPA),] # remove observations without HSGPA
sample <- sample[!is.na(sample$race),] # remove observations without race attribute
sample <- sample[!is.na(sample$sex),] # remove observations without gender attribute
```



## Descriptives

Average time studying and average college GPA

```{r averages}
mean(sample$ClassPrepTime)
mean(sample$OverallGPA)
```

Percentages of female, African-American?
```{r averages2}
mean(sample$female) * 100
mean(sample$africanamerican) * 100
```


## Correlations

What is the relationship between SAT Score and College GPA?

```{r satgpa}
cor(x = sample$SSATScore, y = sample$OverallGPA)
```

What is the relationship between High School GPA and College GPA?
```{r sathsgpa}
cor(x = sample$HSGPA, y = sample$OverallGPA)
```

Which one predicts more of the variance in College GPA?


## Hypothesis test

Is there a difference in SAT scores between men and women?
```{r satdiff}
anova(lm(SSATScore~factor(sex), data=sample)) # anova
summary(lm(SSATScore~factor(sex), data=sample)) # t-test of regression coefficient

```

Is there a difference in College GPA between men and women?
```{r gpadiff}
anova(lm(OverallGPA~factor(sex), data=sample)) # anova
summary(lm(OverallGPA~factor(sex), data=sample)) # t-test of regression coefficient

```


Are there differences in College GPA among members of different racial groups?
```{r gpadiff_race}
anova(lm(OverallGPA~factor(race), data=sample)) # anova
summary(lm(OverallGPA~factor(race), data=sample)) # t-test of regression coefficient

```



## Multiple regression

Controlling for pre-existing ability, are there differences in College GPA among members of different racial groups?
```{r reg1}
anova(lm(OverallGPA~factor(race) + HSGPA, data=sample)) # anova
summary(lm(OverallGPA~factor(race) + HSGPA, data=sample)) # t-test of regression coefficient

```

Does time spent on Facebook predict Overall College GPA when controlling for sex, race, and prior academic ability?
```{r fb}
anova(lm(OverallGPA~factor(race) + factor(sex) + HSGPA + facebookminutesselfreport, data=sample)) # anova
summary(lm(OverallGPA~factor(race) + factor(sex) + HSGPA + facebookminutesselfreport, data=sample)) # t-test of regression coefficient

```

